Combination of field and experimental data together with computational models reveal cognitive mechanisms behind cleaner fish behaviour

### 2022-02-03 manuscript_1.0/manuscript_1.0.tex
Edit's related to Redouan's comments

%%%% Abstract text to be placed here %%%%%%%%%%%%
\begin{abstract}
While it is generally straightforward to quantify individual performance in cognitive experiments, identifying the underlying cognitive processes remains a major challenge. Often, different mechanistic underpinnings may yield similar performances, and Lloyd Morgan's cannon warrants acceptance of the simpler explanation. Alternatively, when the different mechanisms interact with environmental conditions, variation in performance across environments might allow to statistically infer the mechanism responsible. However, to do this it is necessary to have quantitative predictions for the different candidate mechanisms. Here, we use a set of computational models to get quantitative predictions from alternative learning mechanisms, and using Bayesian statistics fit the model parameters to performance data. We used experimental data on performance in an ephemeral reward task by wild-caught cleaner fish Labroides dimidiatus, as well as cleaner and client fish densities from the locations of capture. The task can in principle be solved by estimating future consequences of an action, or by perceiving the removal of the ephemeral reward as psychological punishment (negative reinforcement). We found that a model where cleaners only estimate the future consequences of their actions explains best which cleaner relative abundances cause the fish to develop a preference for an ephemeral food source. This model also yields performances that can be considered the result of locally optimal decision-rules, in contrast to the negative reinforcement model. We argue that the combination of computational models with data is a powerful tool to infer the mechanistic underpinning behind cognitive performance.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Cooperation among individuals of the same or different species is
ubiquitous in nature. Whenever such interactions involve some degree
of partner choice, biological market theory predicts that the exchange
rate of goods and/or services traded adjusts to the law of supply and demand
\citep{noe_Biological_1995a}. While biological market models have largely focused
on evolutionary time scales \citep{noe_Biological_1994a, johnstone_Mutualism_2008a}
\(\color{red}{\text{Hoekstra xxxx}}\), supply and demand conditions, which typically
depend on the abundance of the species involved, certainly vary in time
and space within ecological time scales. Therefore, natural selection should
favour the ability to flexibly adjust decisions and behavioural output to
current market conditions. Indeed, such adjustments have been repeatedly
documented \citep{axen_Signalling_1996}. In animals, an obvious general candidate
mechanism for the strategic adjustment is the cognitive machinery. However,
it is not clear which cognitive mechanisms allow individuals to adjust their
behaviour to the varying market conditions. As adjustments to changes in
market conditions have been documented in a great variety of species with
highly variable brain anatomies, the question arises to what extent mechanisms
beyond basic associative learning may be involved.

One example of strategic adjustment in a biological market is the marine
cleaning mutualism involving the cleaner fish \emph{Labroides dimidiatus} and
`client' fish. Clients offer themselves as food patches for cleaners to
remove ectoparasites, which gives the cleaners calories and the clients
improved health \citep{waldie_LongTerm_2011, ros_Does_2011, triki_Effects_2016}. As cleaners have about 2000
interactions per day \citep{grutter_Relationship_1995}, they regularly face
the situation that two clients seek cleaning simultaneously. In such
conditions, it is important to distinguish between `resident' and
`visitor' client species \citep{bshary_Cleaner_2001a}. The residents have
small home ranges and hence access to only the local cleaner, while the
visitors have large home ranges and hence access to two or more
cleaners. Thus, visitors have the option to switch to another cleaner if
being made to wait, while residents must wait for inspection. Indeed,
visitors have been observed to use their partner choice options in that
way \citep{bshary_Choosy_2002}, which may explain why cleaners give
visitors service priority in a field study in the Red Sea
\citep{bshary_Cleaner_2001a}. More generally, cleaners apparently prefer to
interact with visitors \citep{triki_Marine_2021}. Such a preference may be
linked to a client's value as a food patch; visitors are on average
larger than residents \citep{bshary_Cleaner_2001a}, and size is positively
correlated with parasite load \citep{grutter_Relationship_1995a}. However,
early laboratory experiments showed that leaving causes service
priority independent of client value \citep{bshary_Asymmetric_2002}.
The experiments consisted of asimultaneous two-choice task,
where cleaners had to decide which of two
Plexiglas plates, which offered equal amounts of food, to approach first
and eat mashed prawn of it. One plate was made to behave like a
resident, i.e.~it would remain until the cleaner had eaten off it. The
other plate was made to behave like a visitor, i.e.~it would be removed
if not inspected first. Cleaners learned to prefer the visitor plate and
hence obtained the double amount of food \citep{bshary_Asymmetric_2002}.
Adult cleaner fish outperform various primates as well as rats and
pigeons in this original version of the biological market task
\citep{salwiczek_Adult_2012, zentall_Early_2017}; only African grey
parrots solve this task as well \citep{pepperberg_Can_2014}. As the correct
solution must be learned, and primates are known to readily distinguish
between one and two reward units if presented in more conventional ways
\(\color{red}{\text{ref}}\), what makes the task difficult for most non-cleaners?

Recent research on the cognitive tool kit needed to solve an ephemeral
reward task has used a broad approach as proposed by Shettleworth
\citep{shettleworth_Cognition_2009},
who defined cognition as includes all ways in which animals take in information
through the senses, process, retain and decide to act on it. Indeed, perception
of relevant cues is of major importance for individual performance;
i.e.~what relevant information is encoded in the design of the two
plates (colour versus size; \citep{wismer_Cuebased_2019}), or whether
the information is in the colour of
the food \citep{pretot_Comparative_2021, pretot_Comparing_2016}, and
whether the food is visible or hidden/released only when the subjects
made choices on a touch screen \citep{pretot_Factors_2016}. However, identifying
cues as salient is not the only challenge of the ephemeral reward task,
as revealed by proximate learning models. Such models allow varying the
cognitive tool kit and to evaluate which minimal kit is necessary to solve
the task at hand (e.g. \citep{dubois_Model_2021}). Applied to the ephemeral
reward task, learning models showed that basic reinforcement learning does
not suffice to solve the task \citep{prat_Modelling_2022, quinones_Reinforcement_2019}.
This is particularly so when models assume the more complex natural
situation in which cleaners not only face resident-visitor pairs but also
visitor-visitor and resident-resident pairs, as well only a resident or
only a visitor. To be able to give visitors priority over residents,
cleaners need to be able to assess a client's value separately for the
three possible forms of encounter (alone, paired with a fish with the
same strategic option, paired with a fish with the alternative strategic option)
\citep{quinones_Reinforcement_2019}. The ability to keep such separate scores
for the same stimulus has been termed configurational learning,
or chunking, or segmentation (see references in \citep{prat_Modelling_2022}).

To solve the ephemeral reward task, cleaners also need to account for
the future consequences of current decisions. In the model by
Quiñones et al. \citep{quinones_Reinforcement_2019}, this could be achieved in
two non-mutually exclusive ways: through low temporal discounting of
future effects, also termed `chaining' \citep{enquist_Power_2016}; and/or
through perceiving a visitor client leaving as psychological punishment
(i.e.~as a negative reinforcer). Given that a negative reinforcer is part
of the ubiquitous learning mechanism termed operant conditioning
\citep{thorndike_Animal_1898, skinner_Behavior_1938}, this explanation
could be considered the simpler one. Furthermore, accounting for the
future effects of an action is considered an advanced cognitive
ability \citep{suddendorf_Evolution_2007}. Thus, standard logic of
Lloyd Morgan's cannon demands that operant conditioning as the simpler
explanation is to be accepted by default. Here, we explore the
possibility that fitting a model in which the action of these two
mechanisms, operant conditioning and chaining, depends
on model parameters may reveal which process cleaners actually use.
This is because one process may be better at predicting the observed
variation in cleaner fish performance as a function of ecological conditions.

@@

We used the data from \citep{triki_Biological_2019} to fit the parameters of
the model analysed in \citep{quinones_Reinforcement_2019}. For the bayesian
fitting procedure we used Markov Chain Monte Carlo (MCMC) with a
Metropolis sampler to estimate the posterior distribution of the model
parameters. The data consist a of the number of times that a set of
cleaners chose to feed of a visitor client plate in the context of the
market task. This set of cleaners were collected in different locations
and in each one of those locations environmental variables were
measured. These environmental variables recorded from each location are
used as input for the computational model. Given these inputs, the
computational model produces a time series of choices in the market
task. In the model the preference for the two options changes as the
algorithm learns based on the reward that the two options provide. Thus,
we run the model for 10000 steps and use the last 3000 steps as the
preference developed in the model for visitors given the environmental
variables. We use this preference as the predicted probability in a
binomial distribution to calculate the likelihood that the number of
visitor choices obtained from each cleaner is generated by the model. We
sum over all the cleaners the log-likehoods and use the total for the
acceptance of new parameters in the MCMC.

@@

The model consist of a set individual-based simulations where
individuals face a series of choices between two options. Individuals in
the model represent cleaners, and the options they face simulate the
natural conditions of a cleaning market. Thus, individuals in the model
choose to prioritize different clients. These clients are classified as
resident and visitors. Residents are those clients that stay in the
cleaning queue when they are not given priority; while visitors leave
the queue (with a certain probability) after they are not given
priority. Individuals obtain a fixed reward from cleaning a client
regardless of the type. Every time individuals face and make a choice
they update the probability of making that same choice. The update is
based on the difference between the expectation of reward and the actual
reward the individual obtains (prediction error)
\citep{sutton_Reinforcement_2018, rescorla_Theory_1972}; and it is carried
in the direction that would lead to more reward being obtained, given
the new information. Further details of the model implementation can be
found in the paper \citep{quinones_Reinforcement_2019}. The model analysis
points to the relevance of two parameters. First, \(\gamma\) measures how
much individuals include future rewards in their decision updates. If
\(\gamma=0\), individuals only use the immediate reward obtained from a
cleaning interaction. As \(\gamma\) increases individuals include more the
reward obtained from the subsequent choices. That amounts to calculating
and using for decision making the future expected rewards in action.
Second, \(\eta\) measures how much individuals include in their reward the
fleeing behaviour of visitor clients as a negative component. Both of
these parameters allow individuals to use in their estimates the future
effects of their choices. The model analysis by Quiñones \emph{et al}
\citep{quinones_Reinforcement_2019} shows that without either of them, learning
individuals cannot develop a preference for the visitor client. The
analysis contrast the effect of these two parameters against variation
in the relative abundance of the two client types. The analysis shows
that under very high client (relative) abundance the presence of future
rewards (\(\gamma>0\)) no longer favours the preference for the ephemeral
clients, this effect does not occur with negative reward mediated by
client behaviour (\(\eta\)). Thus, the analysis suggests that future
reward as an explanation for cleaner ability to prefer visitors is
supported by Triki \emph{et al} \citep{triki_Biological_2019} data. However, a combination of
future and negative rewards, could still be a better explanation for the
data. In the current analysis, we will use the data from
\citep{triki_Biological_2019} to fit some of the model parameters, particularly
relevant we will fit \(\gamma\) and \(\eta\) to find out whether each or a
combination of these effects is a better explanation for the pattern
seen in the data. For other parameter values used for the model
simulations and the MCMC see Table \ref{tab:param}

\hypertarget{data-explanation}{%
\subsubsection{Data explanation}\label{data-explanation}}

\emph{Insert here brief description of the data and data collection}

In order to capture with the model the relationship between the
characteristics of the market (environment) and the cleaner's
performance we need to translate the absolute abundances of the data to
to a measure of relative abundances that captures client visitation patterns.
This is because in the model, relative abundances
of clients define not only the probability of residents and visitors,
but also how often the cleaning station is empty (e.g.~there are no
clients to be cleaned). The more cleaners there are, the more options
are available for the clients and hence the higher is the probability of
empty cleaning stations. How often the cleaning stations are empty in
turns influences the value that cleaners can expect from their choices.
The emptier the cleaning stations are, the more beneficial it is to
prioritize a client that will not stay. However, the frequency of empty
spots in the cleaning station does not only depend on cleaner abundances
but also on the frequency with which clients visit the
station. The client visitation frequency will likely vary among
different clients depending on parasite abundances and loads. These are
variables for which we do not have field estimates. Thus, we computed a
measure of relative cleaner abundance for each site relative to absolute
abundances and multiplied by a scaling constant that changes the range
of the variable. This scaling constant is meant to capture variation in the market
conditions driven partly by cleaner abundance. We fit the value of the
scaling constant as part of the
statistical inference. As for the client abundances for residents and
visitors we computed a relative measure with respect to the total client
abundance, and weighted that by the rescaled relative absence of
cleaners (1-relative cleaner abundance). Thus, all three measures of
relative abundance sum up to 1, and can be used in the model as a proxy
for the probability of having different options in the cleaning station.

>>>>

%%%% Abstract text to be placed here %%%%%%%%%%%%
\begin{abstract}
While it is generally straightforward to quantify individual performance in cognitive experiments, identifying the underlying cognitive processes remains a major challenge. Often, different mechanistic underpinnings may yield similar performances, and Lloyd Morgan's cannon warrants acceptance of the simpler explanation. Alternatively, when the different mechanisms interact with environmental conditions, variation in performance across environments might allow to statistically infer the mechanism responsible. However, to do this it is necessary to have quantitative predictions for the different candidate mechanisms. Here, we use a set of computational models to get quantitative predictions from alternative learning mechanisms, and using Bayesian statistics fit the model parameters to performance data. We used experimental data on performance in an ephemeral reward task by wild-caught cleaner fish \emph{Labroides dimidiatus}, as well as cleaner and client fish densities from the locations of capture. The task can in principle be solved by estimating future consequences of an action, or by perceiving the removal of the ephemeral reward as psychological punishment (negative reinforcement). We found that a model where cleaners only estimate the future consequences of their actions explains best which cleaner relative abundances cause the fish to develop a preference for an ephemeral food source. This model also yields performances that can be considered the result of locally optimal decision-rules, in contrast to the negative reinforcement model. We argue that the combination of computational models with data is a powerful tool to infer the mechanistic underpinning behind cognitive performance.
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Cooperation among individuals of the same or different species is
ubiquitous in nature. Whenever such interactions involve some degree
of partner choice, biological market theory predicts that the exchange
rate of goods and/or services traded adjusts to the law of supply and demand
\citep{noe_Biological_1995a}. While biological market models have largely focused
on evolutionary time scales \citep{noe_Biological_1994a, johnstone_Mutualism_2008a}
\(\color{red}{\text{Hoekstra xxxx}}\), supply and demand conditions, which typically
depend on the abundance of the species involved, certainly vary in time
and space. Therefore, natural selection should
favour the ability to flexibly adjust decisions and behavioural output to
current market conditions. Indeed, such adjustments have been repeatedly
documented \citep{axen_Signalling_1996}. In animals, an obvious general candidate
mechanism for the strategic adjustment is the cognitive machinery. However,
it is not clear which cognitive mechanisms allow individuals to adjust their
behaviour to the varying market conditions. As adjustments to changes in
market conditions have been documented in a great variety of species with
highly variable brain anatomies, the question arises to what extent mechanisms
beyond basic associative learning may be involved.

One example of strategic adjustment in a biological market is the marine
cleaning mutualism involving the cleaner fish \emph{Labroides dimidiatus} and
`client' fish. Clients offer themselves as food patches for cleaners to
remove ectoparasites, which gives the cleaners calories and the clients
improved health \citep{waldie_LongTerm_2011, ros_Does_2011, triki_Effects_2016}. As cleaners have about 2000
interactions per day \citep{grutter_Relationship_1995}, they regularly face
the situation that two clients seek cleaning simultaneously. In such
conditions, it is important to distinguish between `resident' and
`visitor' client species \citep{bshary_Cleaner_2001a}. The residents have
small home ranges and hence access to only the local cleaner, while the
visitors have large home ranges and hence access to two or more
cleaners. Thus, visitors have the option to switch to another cleaner if
being made to wait, while residents must wait for inspection. Indeed,
visitors have been observed to use their partner choice options in that
way \citep{bshary_Choosy_2002}, which may explain why cleaners give
visitors service priority in a field study in the Red Sea
\citep{bshary_Cleaner_2001a}. More generally, cleaners apparently prefer to
interact with visitors \citep{triki_Marine_2021}. Such a preference may be
linked to a client's value as a food patch; visitors are on average
larger than residents \citep{bshary_Cleaner_2001a}, and size is positively
correlated with parasite load \citep{grutter_Relationship_1995}. However,
early laboratory experiments showed that leaving causes service
priority independent of client value \citep{bshary_Asymmetric_2002}.
The experiments consisted of asimultaneous two-choice task,
where cleaners had to decide which of two
Plexiglas plates, which offered equal amounts of food, to approach first
and eat mashed prawn of it. One plate was made to behave like a
resident, i.e.~it would remain until the cleaner had eaten off it. The
other plate was made to behave like a visitor, i.e.~it would be removed
if not inspected first. Cleaners learned to prefer the visitor plate and
hence obtained the double amount of food \citep{bshary_Asymmetric_2002}.
Adult cleaner fish outperform various primates as well as rats and
pigeons in this original version of the biological market task
\citep{salwiczek_Adult_2012, zentall_Early_2017}; only African grey
parrots solve this task as well \citep{pepperberg_Can_2014}. As the correct
solution must be learned, and primates are known to readily distinguish
between one and two reward units if presented in more conventional ways
\(\color{red}{\text{ref}}\), what makes the task difficult for most non-cleaners?

Recent research on the cognitive tool kit needed to solve an ephemeral
reward task has used a broad approach as proposed by Shettleworth
\citep{shettleworth_Cognition_2009},
who defined cognition as including all ways in which animals take in information
through the senses, process, retain and decide to act on it. Indeed, perception
of relevant cues is of major importance for individual performance;
i.e.~what relevant information is encoded in the design of the two
plates (colour versus size; \citep{wismer_Cuebased_2019}), or whether
the information is in the colour of
the food \citep{pretot_Comparative_2021, pretot_Comparing_2016}, and
whether the food is visible or hidden/released only when the subjects
made choices on a touch screen \citep{pretot_Factors_2016}. However, identifying
cues as salient is not the only challenge of the ephemeral reward task,
as revealed by proximate learning models. Such models allow varying the
cognitive tool kit and to evaluate which minimal kit is necessary to solve
the task at hand (e.g. \citep{dubois_Model_2021}). Applied to the ephemeral
reward task, learning models showed that basic reinforcement learning does
not suffice to solve the task \citep{prat_Modelling_2022, quinones_Reinforcement_2019}.
This is particularly so when models assume the more complex natural
situation in which cleaners not only face resident-visitor pairs but also
visitor-visitor and resident-resident pairs, as well only a resident or
only a visitor. To be able to give visitors priority over residents,
cleaners need to be able to assess a client's value separately for the
three possible forms of encounter (alone, paired with a fish with the
same strategic option, paired with a fish with the alternative strategic option)
\citep{quinones_Reinforcement_2019}. The ability to keep such separate scores
for the same stimulus has been termed configurational learning,
or chunking, or segmentation (see references in \citep{prat_Modelling_2022}).

To solve the ephemeral reward task, cleaners also need to account for
the future consequences of current decisions. In the model by
Quiñones \emph{et al.} \citep{quinones_Reinforcement_2019}, this could be achieved in
two non-mutually exclusive ways: through low temporal discounting of
future effects, also termed `chaining' \citep{enquist_Power_2016}; and/or
through perceiving a visitor client leaving as psychological punishment
(i.e.~as a negative reinforcer). Given that a negative reinforcer is part
of the ubiquitous learning mechanism termed operant conditioning
\citep{thorndike_Animal_1898, skinner_Behavior_1938}, this explanation
could be considered the simpler one. Furthermore, accounting for the
future effects of an action is considered an advanced cognitive
ability \citep{suddendorf_Evolution_2007}. Thus, standard logic of
Lloyd Morgan's cannon demands that operant conditioning as the simpler
explanation is to be accepted by default. Here, we explore the
possibility that fitting a model in which the action of these two
mechanisms, operant conditioning and chaining, depends
on model parameters may reveal which process cleaners actually use.
This is because one process may be better at predicting the observed
variation in cleaner fish performance as a function of ecological conditions.

@@

To test the relative validity of chaining versus negative reinforcement as
mechanisms to account for future consequences of current choices,
we combined empirical data from Triki \emph{et. al}
\citep{triki_Biological_2019, triki_Brain_2020}. with the computational
model analysed in Quiñones \emph{et al.}\citep{quinones_Reinforcement_2019}. In order to
compare the explanatory power of the different hypothesis, we estimate the
parameter values of three versions of the model that best fit the data.
Particularly, we fit the parameter that determines how important
future rewards are (\(\gamma\)), the magnitude of the negative reward
(\(\eta\)) and a scaling constant that translates absolute to relative
abundances (see following sections for details). We obtained model
predictions using a set of stochastic individual-based
simulations. Thus, each parameter combination of the model yields not one
prediction for the data set, but a distribution of them. Thus, for the
fitting procedure, we use one of the most common algorithms in bayesian inference,
the Markov Chain Monte Carlo (MCMC). MCMC is a stochastic algorithm that allows
to sample from the posterior distribution of model parameter values. The posterior
distribution represents the most likely parameter values given the data.
Due to the stochastic nature of MCMC, which repeatedly samples parameter values
and evaluates their likelihood, the distribution of likelihoods from the
stochastic model is also explored. This makes MCMC and bayesian inference
particularly suitable for the computational models we evaluate. Bellow,
we explain first the computational model and the data collection, and later
how the MCMC was used to estimate parameter values.

@@

The model consist of a set individual-based simulations where
individuals face a series of choices between two options. Individuals in
the model represent cleaners, and the options they face simulate the
natural conditions of a cleaning market. Cleaners experience a series
of discrete time points in which they face different `states', which
are defined by the number and type of clients inviting for service.
There are six possible states: zero clients, one resident, one visitor,
two residents, two visitors, and a mix of one resident and one visitor.
The probability of each state is largely determined by the relative
abundance of cleaners, residents and visitors, but to some degree by a
cleaner's choices when it faces the resident-visitor combination.
This is because residents are those clients that stay in the
cleaning queue when they are not given priority and hence would
still be available in the next time point; while visitors leave
the queue (with a certain probability) after they are not given
priority. Individuals obtain a fixed reward from cleaning a client
regardless of the type. Every time individuals face and make a choice
they update the probability of making that same choice. The update is
based on the difference between the expectation of reward and the actual
reward the individual obtains (prediction error)
\citep{sutton_Reinforcement_2018, rescorla_Theory_1972}; and it is carried
in the direction that would lead to more reward being obtained, given
the new information. In the long run, the probability of choosing a visitor
over a resident converges in the model. To which probability the model
will converge depends on the relative abundance of cleaners, visitors
and resident; as well as on the probability of visitors leaving the station
when unattended. Further details of the model implementation can be
found in the paper \citep{quinones_Reinforcement_2019}.

The model shows that only agents that calculate rewards separately for
the different states (termed `fully aware agents') can potentially
prioritise visitors of residents in nature, and hence also solve the
ephemeral reward task. This is because pooling rewards for the two
clients from all states by `partially aware agents'
(that only recognize the two clients as different) cancels out
all differences (see \citep{quinones_Reinforcement_2019}). However,
such \emph{chunking} by fully aware agents is not enough to develop a
preference for visitors. In addition, agents need to find a mean
to incorporate future consequences of current choices. In the model,
this could be achieved with either of two parameters that could
also work together. First, \(\gamma\) measures how
much individuals include future rewards in their decision updates. If
\(\gamma=0\), individuals only use the immediate reward obtained from a
cleaning interaction. As \(\gamma\) increases individuals include more the
reward obtained from the subsequent choices. That amounts to calculating
and using for decision making the future expected rewards of an action.
Second, \(\eta\) measures how much individuals include in their reward the
fleeing behaviour of visitor clients as a negative component. Both of
these parameters allow individuals to use in their estimates the future
effects of their choices.

A further model analysis contrast the effect of these two
parameters against variation in the relative abundance of cleaners
and the two client types. Assuming certainty that visitors leave, the
two parameters yield different
predictions when relative client abundance is very high. Under such
conditions, the analysis shows that under very high client (relative)
abundance the presence of future rewards (\(\gamma>0\)) no longer favours
the preference for the ephemeral clients, while the preference persists
when the negative reward mediated by client behaviour (\(\eta>0\)) is
included. Thus, the analysis suggests that future reward as an
explanation for cleaner ability to prefer visitors is supported.
At first sight, empirical data seem to support future rewards as
the mechanism used by cleaners, as the statistical analysis by
Triki \emph{et al.} \citep{triki_Biological_2019} data revealed that low
cleaner-to-client ratios predict low preference for visitors and hence
low performance in the ephemeral reward task. However, low cleaner-to-client
ratios may also lead to low probabilities of visitors leaving, which would
cause low preferences for visitors independently of the underlying
mechanism \citep{quinones_Reinforcement_2019}. Furthermore, a combination
of future and negative rewards, could still be yield a better explanation
for the data than any single factor.

\hypertarget{empirical-data}{%
\subsubsection{Empirical data}\label{empirical-data}}

The empirical data were collected between 2010 and 2018 always during
the austral winter months June to August at Lizard Island
(\(14.6682° S, 145.4604° E\)), Great Barrier Reef, Australia.
They consist of three sets: fish densities from a total of five study
reef sites (Corner Beach-CB, Horseshoe-HS,
Mermaid Cove-MC, Northern Horseshoe-NHS, and The Crest-TC),
natural observations of cleaners to quantify the probability
of visitors leaving if initially ignored in favour of another client, and the
performance of individuals caught from these sites in the ephemeral reward test.
In total, we have twelve site/year data sets for fish densities and
corresponding performance in the test. Thus, some sites were sampled more
than once. Due to major environmental perturbations during the study period
(\(\color{red}{\text{refs}}\)), fish densities changed significantly even
within sites \citep{triki_Decrease_2018, triki_Fluctuations_2019}. To estimate
the population density of cleaner fish and their clients at a given site in
a given year, Triki \emph{et al.} \citep{triki_Biological_2019} used a series of ten
transects of \(30m\) each. Observers swam along the transect lines placed on
the reef and first counted the visible large-bodied adult fish
(species with total length TL \textgreater{} \(10cm\)) on a width of \(5m\), and then on the
return individuals of small-bodied fish species (TL \textless{} \(10 cm\)) on a width
of \(1 m\). We included cleaner fish in the \(5 m\) transect. Based on the
classification by Bshary \citep{bshary_Cleaner_2001a} and Wismer \emph{et al.}
\citep{wismer_Variation_2014}, all visitor client species (\(n = 124\)) were
large-bodied. Resident clients comprised all small-bodied species
(\(n = 30\)) but also some large-bodied ones (\(n = 15\)). Overall,
large-bodied species are highly correlated with being a visitor,
and small-bodied species are a good correlate of residents. Based on the
transect data, Triki \emph{et al.} \citep{triki_Biological_2019, triki_Brain_2020}
estimated cleaner fish, small-bodied client, large-bodied client and total
client fish densities per \(100 m^2\). LASSO analyses revealed that cleaner and
large-bodied client densities were highly correlated, and only the former
was hence kept in the analyses as representative of both \citep{triki_Biological_2019}.
Cleaner densities, in interaction with visitor client leaving probability,
explained best individual performance in the ephemeral reward task
\citep{triki_Biological_2019}, while cleaner-to-client ratio was not a significant
factor. This creates a certain mismatch with the model by Quiñones \emph{et al.}
\citep{quinones_Reinforcement_2019}, which focused on abundance of cleaners relative
to overall client abundance in order to capture the characteristics
of the market. Importantly, relative abundances of clients define not only
the probability of interactions with residents and visitors, but also affect
how often the cleaning station is empty (e.g.~there are no clients to be cleaned).
The higher the relative abundance of cleaners is, the higher is the probability
of cleaning stations being empty at a given time point. How often the
cleaning stations are empty in turns influences the value that cleaners can
expect from their choices. The emptier the cleaning stations are, the more
beneficial it is to prioritize a client that will not stay. On the other hand,
situations where cleaners have to choose will be rare if cleaner relative
abundance is high, which reduces the total value of prioritizing visitors
over residents and makes it more difficult to learn a preference for visitors.
Given the model structure, we used the empirical data to calculate relative
abundances of cleaners, residents and visitors for the current analyses.
To incorporate the empirical finding that absolute densities matter, we
introduced a new scaling constant (see section on model fitting).

To collect data on the visitor clients' probability of leaving if not serviced
immediately, eight cleaners per site/year were each filmed for \(30 min\).
From the videos, all events in which a visitor client was made to wait in
favour of another (visitor or resident) client were identified, and noted
whether or not the visitor left \citep{triki_Biological_2019, triki_Brain_2020}.
A total of 120 cleaners (10 individuals per 12 site/year) were tested in the
ephemeral reward task \citep{triki_Biological_2019, triki_Brain_2020}. Authors
housed all captured cleaners individually in glass aquaria
( \(62cm \times 27cm \times 37 cm\) ) and provided them
with PVC pipes (\(10 cm \times 1 cm\)) as shelters.
The task consisted of exposing the cleaner fish to substitute
models of client fish in the form of two \emph{Plexiglas} plates offering the
same amount of food (one item of mashed prawn). The two plates differed
in colour and pattern (horizontal green stripes or vertical pink stripes)
but had equal size (\(10 cm \times 7 cm\)). Importantly, the two plates played
different roles as either a visitor (ephemeral food source) or
resident (permanent food source). That is, if a cleaner inspected the
resident plate first, the experimenter withdrew the visitor plate out of
the aquarium as a consequence. Choosing first the visitor plate,
however, granted access to both plates. The equal size of the plates
forced cleaner fish to learn to give service priority to the visitor plate
based solely on the behaviour-cue of the plates rather than size-cue
\citep{wismer_Cuebased_2019}. Triki \emph{et al.} \citep{triki_Biological_2019, triki_Brain_2020}
tested the fish for a maximum of 200 trials with 20 trials a day, 10 trials
in the morning and 10 trials in the afternoon. They randomized and
counterbalanced the plates' spatial location (i.e.~left or right)
between trials. Similarly, they counterbalanced the plates' decoration
(colour and pattern) and the plates' role (visitor or resident) between the
tested fish. The data was originally collected to assess which variables are
good predictors of the capacity to solve the ephemeral food task. Thus, previous
analysis use specific statistical criteria to classify individual cleaners'
performance as successful or failure \citep{triki_Biological_2019}. Because the model
predicts the preference of cleaners given the set of environmental variable,
we instead use the raw data of individuals cleaners' decisions.
We use 20 choices from each cleaner as the preference they develop
for the visitor plate. Which these 20 choices were vary for each cleaner depending
on the performance that was originally assessed. For cleaners that did not
succeed, the 20 choices correspond to the last two sessions. Cleaners that
succeeded in solving the task were tested on a reversal task. That is, the
pattern in plates was switched to the opposite role they played in the
initial test. For those cleaners, we used the last session before they succeeded
in the initial, and the last session of the reversal, completing the 20 choices.
We did not distinguish between individuals that reached the criterion early
or late during the 200 trials.

\hypertarget{statistical-analysis}{%
\subsubsection{Statistical analysis}\label{statistical-analysis}}

The aim of the analysis is to fit the key model parameters that allow
individuals to incorporate future consequences of current actions,
\(\gamma\) and \(\eta\), to the data from Triki \emph{et al.}
\citep{triki_Biological_2019, triki_Brain_2020} to find out whether each or
a combination of these effects is a better explanation for the pattern
seen in the data. In particular, we are interested in the relation between the
ecological variables: cleaners, visitor clients, resident clients abundances and
visitor clients leaving probability; and cleaners' preference for visitor clients
over residents. Thus, the four ecological variables are given as input to
the models. The preference for visitors gotten from the simulations of the model
is used as the prediction for the response variable, while the number of visitors
chosen out of the 20 choices of each cleaner as the observed response variable.
We kept all other parameter values used for the model simulations constant,
see Table \ref{tab:param}.

In order to capture with the model the relationship between the characteristics
of the market (environment) and the cleaner's preference we need to translate
the absolute abundances of the data to to a measure of relative abundances
that captures client visitation patterns. This is because in the model, relative
abundances of clients define not only the probability of residents and visitors,
but also how often the cleaning station is empty (e.g.~there are no
clients to be cleaned). The frequency with which clients visit the station
is another variable, which will likely vary among different client species
depending on parasite abundances and loads. We do not have field estimates
for species specific parasite loads, especially not as a function of site.
In order to control for these aspects, we computed a measure of relative
cleaner abundance for each site relative to absolute abundances and multiplied
by a scaling constant that changes the range of the variable. This scaling
constant is hence meant to capture variation in the market conditions driven
partly by cleaner abundance. We fit the value of the scaling constant as
part of the statistical inference. As for the client abundances for
residents and visitors we computed a relative measure with respect to the
total client abundance, and weighted that by the rescaled relative absence
of cleaners (\(1-\text{relative cleaner abundance}\)). Thus, all three measures
of relative abundance sum up to 1, and can be used in the model as a proxy
for the probability of having different options in the cleaning station.
Note that the scaling constant was introduced to control for variation that
is not captured by the model; its parameter distribution does not offer
biological insights.

Once we calculate the relative abundances, we can obtain predictions from
the model for each one of the locations and run the Markov Chain. We start the
chain with random values for the parameters of interest, then run the
computational model once for every location using as input the ecological
explanatory variables. The model outputs the probability of choosing the visitor
for each location \(p_i\), where \(i\) is the index for the 12 locations. Assuming a
binomial distribution, the probability that each of the cleaners' 20 choices was
generated by the model is given by \(\binom{n}{k_ij}p^{k_{ij}}_i (1-p_i)^{n-k_{ij}}\);
where \(k_{ij}\) is the number of times that cleaner \(j\) in location \(i\) chose the
visitor over the resident; and \(n=20\) (due to our choice of using 20
choices per cleaner). By taking the natural logarithm and summing over all
the cleaners and locations we obtain the log-likelihood of the data
given the model and parameters. We then propose a new set of parameters drawn
from a uniform distribution centred around the old parameter set. The amplitude
of the uniform distribution used for each parameter can be found in Table
\ref{tab:param}. Subsequently, we ran the model and calculated the likelihood
with the new parameter set. We then use the ratio of the two likelihoods to
choose which parameter set to keep. New parameter sets with a higher likelihood
than the old set replace old ones, and those with a lower likelihood are replace
current ones with a probability equals to the log-likelihood ratio.
Given that we only use the likelihoods in the decision, we are using a
uninformative prior. Once we decide whether the new parameter set replaces
the old one, we ran the model again to sample to the likelihood
distribution of the parameter set. We then start
the cycle again by proposing a new set of parameters, and repeat the process for
\(1e^5\) steps. We ran 5 independent chains, discarded the first
\(100\) samples of each chain as burn in, and after that we keep 1 in every
100 samples to avoid autocorrelation. The collection of parameters sets kept in
all chains approximates the posterior distribution. The model as well as the
fitting algorithm were coded in \emph{c++}, the diagnostics and visualization were
coded in R \citep{rcoreteam_Language_2021} of the all code can be found at
\(\color{red}{\text{include citation}}\)

In order to compare the fit that the three alternative models give we use the
distribution of pseudo \(R^2\) proposed by Mcfadden \citep{mcfadden_Conditional_1974}.
Mcfadden's pseudo \(R^2\) is a standard measure of fit for logistic regression.
In that context, \(pseudo-R^2\) uses the log-likelihood of the data given
the model, relative to the log-likelihood of the data given a
model without covariates, as a measure of fit. Given that our
model is not a logistic regression, we measure the \(pseudo-R^2\) in
relation to the log-likelihood of a model
with parameters \(\gamma\) and \(\eta\) set to zero. This in practice
amounts to a model that has neutral preference between the two options.
Pseudo \(R^2\) was calculated for all the samples of the posterior gotten from
the MCMC. Thus, we use these distributions of pseudo \(R^2\)'s as a measure of
fit.
